Learning a reward function from demonstrations: a cascaded supervised learning approach

### 2012-10-31 piot2012cascading/ECML13_cascading.tex
corrections de fomre mineures

% Created 2012-10-15 lun. 14:49
\documentclass[smallextended]{svjour3}
\smartqed 
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage[integrals]{wasysym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subfigure}
\tolerance=1000
\usepackage{amsmath}
%\usepackage{amsthm}
%\newtheorem{definition}{Definition}
%\newtheorem{theorem}{Theorem}
\newtheorem{cor}{Corollary}
\usepackage{dsfont}
\usepackage{mathtools}
\mathtoolsset{showonlyrefs=true}
\providecommand{\alert}[1]{\textbf{#1}}
\newcommand{\E}{\mathbf{E}}
\newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
\newcommand{\Card}{\operatorname*{Card}} %\operatorname* pour les op. pouvant admettre des limites...
\begin{document}
\title{Learning a reward function from demonstrations: a cascaded supervised learning approach}
\author{Edouard Klein$^{1,2}$ \and Bilal Piot $^{2,3}$\and Matthieu Geist $^{2}$\and Olivier Pietquin$^{2,3}$}
\titlerunning{Cascading approach to IRL}
\institute{
 1. ABC Team\\
 LORIA-CNRS, France.\\
\and
2.Supélec\\
 IMS-MaLIS Research group, France\\
 \texttt{firstname.lastname@supelec.fr}\\
\and 
3. UMI 2958 GeorgiaTech-CNRS\\
France\\
}
\date{\today}

@@

Having in mind that there is a one to one relation between a reward function and its associated optimal action-value function (via the Bellman equation, see Eq.~\ref{eq:bellman1}), it is worth thinking of a method able to output an action-value function for which the greedy policy is the demonstrated expert policy. Thus, the demonstrated expert policy will be optimal for the related reward function. We propose a score function-based multi-class classification step (see Sec.~\ref{sec:algo}) to infer a score function. Besides, in order to retrieve via the Bellman equation the reward associated to the score function computed by the classification step, we introduce a regression step (see Sec.~\ref{sec:algo}). That is why the method is called the Cascaded Supervised Inverse reinforcement learning (CSI). This method is analyzed, see Sec.~\ref{sec:analysis} where it is shown that the demonstrated expert policy is near-optimal for the reward the regression step outputs.

@@

Given a dataset $D_C = \{(s_i,a_i=\pi^E(s_i))_i\}$ of actions $a_i$ (deterministically) chosen by the expert on states $s_i$, we train such a classifier. The classification policy $\pi_C$ is not the end product we are looking for (that would be mere imitation of the expert, not IRL). What is of particular interest to us is the score function $q$ itself. One can easily notice the similarity between Eq.~\ref{eq:greedy2} and Eq.~\ref{eq:greedy} that describes the relation between the optimal policy in an MDP and its optimal action-value function. It is only natural to view the score function $q$ of the classifier as some kind of optimal action-value function for the classifier policy $\pi_C$. By inversing the Bellman equation Eq.~\ref{eq:bellman1} with $q$ in lieu of $Q^\pi_R$, one gets $R^C$, the reward function relative to our score/action-value function $q$:
\begin{equation}
  \label{eq:rc}
  R^C(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi_C(s')).
\end{equation}
As we wish to solve the general, or approximate, IRL problem where the transition probabilities $P$ are unknown, our reward function $R^C$ will be approximated with the help of information gathered by interacting with the system. We assume that another dataset $D_R = \{(s_j,a_j,s'_j)_j\}$ is available where $s'_j$ is the state an agent taking action $a_j$ in state $s_j$ transitionned to. Action $a_j$ needs not be chosen by any particular policy. The dataset $D_R$ brings us information about the dynamics of the system. From it, we construct datapoints
\begin{equation}
  \label{eq:rj}
  \{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j.
\end{equation}
As $s'_j$ is sampled accordingly to $P(\cdot|s_j,a_j)$ the constructed datapoints help building a good approximation of $R^C(s=s_j,a=a_j)$. A regressor (a simple least-square approximator can do but other solutions such as Gaussian Processes \cite{rasmussen2006gaussian} could also be envisionned) is then fed the datapoints $((s_j,a_i),\hat r_j)$ to obtain $\hat R^C$, a generalisation of $\{((s_j,a_j),\hat r_j)_j\}$ over the whole state-action space. The complete algorithm is given in Alg.~\ref{algo:cascading}.
\begin{algorithm}%[H]
    %\small
  \caption{CSI algorithm}
  \label{algo:cascading}
  \emph{\textbf{Given}} a training set $D_C=\{(s_i,a_i=\pi^E(s_i))\}_{1\leq i \leq D}$ and another training set $D_R=\{(s_{j},a_{j},s'_{j})\}_{1\leq j \leq D'}$\;\\
  \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$\;\\
  \emph{\textbf{Learn}} a reward function $\hat R^C$ from the dataset $\{((s_{j},a_{j}),\hat{r}_j)\}_{1\leq j \leq D'}$, $\forall (s_j,a_j,s'_j) \in D_R,\hat{r}_j=q(s_{j},a_{j})-\gamma q(s'_{j},\pi_C(s'_{j}))$\;\\
  \emph{\textbf{Output}} the reward function $\hat R^{C}$ \;
\end{algorithm}

@@

  The function $q$ is the action-value function for $\pi_C$ with respect to the reward $R^C$, by definition (see Eq.~\ref{eq:rc}).

@@

More precisely, the expert controls the system for $n$ runs of length $n$, with values of $n$ in $\{3,7,10,15,20\}$, giving $n^2$ samples $(s_k,a_k=\pi_E(a_k),s'_k)_k = D_E$. The dataset $D_C$ fed to the classifier is straightforwardly constructed from $D_E$ by dropping the $s'_k$ terms:
\begin{equation}
  D_C = \{(s_i = s_k,a_i=a_k)_i\}.
\end{equation}
There is no particuliar problem for the classification step. It is however not reasonable to construct the dataset $D_R$ only from expert transitions and expect a small regression error term $\epsilon_R$. Indeed, the dataset $D_E$ only sample the dynamics induced by the expert's policy and not the whole dynamics of the MDP. The regression step also needs samples $((s_j,a_j),\hat r_j)_j$ drawn from non expert actions. We address this problem by using a heuristics.
\subsubsection{Heuristics}
\label{subsubsec:heuristics}
From the raw expert dataset $D_E$ we can construct samples $((s_j=s_k,a_j=a_k),\hat r_j)$ with $\hat r_j$ as specified in Eq.~\ref{eq:rj}. To compensate for the lack of information about the dynamics of the whole MDP, we assert that disagreeing with the expert's choice is undesirable. Mathematically, we do that by artificially introducting samples $((s_j=s_k,a),r_{min})_{j;\forall a\neq \pi_E(s_j) = a_k}$ where
\begin{equation}
  r_{min} = \min_j\hat r_j - 1.
\end{equation}
Although this heuristics was not analyzed in Sec.~ \ref{sec:analysis} (where the availability of a more complete dataset $D_R$ was assumed), the results shown in the next two subsections demonstrate its soundness.
\subsubsection{Subroutines choices}
\label{subsubsec:subroutines}
The cascading algorithm can be instanciated with some standard classification algorithms and any regression algorithm. The choice of such subroutines may be dictated by the kind and amount of available data, by ease of use or by computational complexity, for example.

>>>>

% Created 2012-10-15 lun. 14:49
\documentclass[smallextended]{svjour3}
\smartqed 
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage[integrals]{wasysym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subfigure}
\tolerance=1000
%\usepackage{amsmath}
%\usepackage{amsthm}
%\newtheorem{definition}{Definition}
%\newtheorem{theorem}{Theorem}
\newtheorem{cor}{Corollary}
\usepackage{dsfont}
\usepackage{mathtools}
\mathtoolsset{showonlyrefs=true}
\providecommand{\alert}[1]{\textbf{#1}}
\newcommand{\E}{\mathbf{E}}
\newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
\newcommand{\Card}{\operatorname*{Card}} %\operatorname* pour les op. pouvant admettre des limites...
\begin{document}
\title{Learning a reward function from demonstrations: a cascaded supervised learning approach}
\author{Edouard Klein$^{1,2}$ \and Bilal Piot $^{2,3}$\and Matthieu Geist $^{2}$\and Olivier Pietquin$^{2,3}$}
\titlerunning{Cascading approach to IRL}
\institute{
 1. ABC Team\\
 LORIA-CNRS, France.\\
\and
2.Supélec\\
 IMS-MaLIS Research group, France\\
 \texttt{firstname.lastname@supelec.fr}\\
\and 
3. UMI 2958 GeorgiaTech-CNRS\\
France\\
}
\date{\today}

@@

Having in mind that there is a one to one relation between a reward function and its associated optimal action-value function (via the Bellman equation, see Eq.~\eqref{eq:bellman1}), it is worth thinking of a method able to output an action-value function for which the greedy policy is the demonstrated expert policy. Thus, the demonstrated expert policy will be optimal for the related reward function. We propose a score function-based multi-class classification step (see Sec.~\ref{sec:algo}) to infer a score function. Besides, in order to retrieve via the Bellman equation the reward associated to the score function computed by the classification step, we introduce a regression step (see Sec.~\ref{sec:algo}). That is why the method is called the Cascaded Supervised Inverse reinforcement learning (CSI). This method is analyzed, see Sec.~\ref{sec:analysis} where it is shown that the demonstrated expert policy is near-optimal for the reward the regression step outputs.

@@

Given a dataset $D_C = \{(s_i,a_i=\pi^E(s_i))_i\}$ of actions $a_i$ (deterministically) chosen by the expert on states $s_i$, we train such a classifier. The classification policy $\pi_C$ is not the end product we are looking for (that would be mere imitation of the expert, not IRL). What is of particular interest to us is the score function $q$ itself. One can easily notice the similarity between Eq.~\eqref{eq:greedy2} and Eq.~\eqref{eq:greedy} that describes the relation between the optimal policy in an MDP and its optimal action-value function. It is only natural to view the score function $q$ of the classifier as some kind of optimal action-value function for the classifier policy $\pi_C$. By inversing the Bellman equation Eq.~\eqref{eq:bellman1} with $q$ in lieu of $Q^\pi_R$, one gets $R^C$, the reward function relative to our score/action-value function $q$:
\begin{equation}
  \label{eq:rc}
  R^C(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi_C(s')).
\end{equation}
As we wish to solve the general, or approximate, IRL problem where the transition probabilities $P$ are unknown, our reward function $R^C$ will be approximated with the help of information gathered by interacting with the system. We assume that another dataset $D_R = \{(s_j,a_j,s'_j)_j\}$ is available where $s'_j$ is the state an agent taking action $a_j$ in state $s_j$ transitionned to. Action $a_j$ needs not be chosen by any particular policy. The dataset $D_R$ brings us information about the dynamics of the system. From it, we construct datapoints
\begin{equation}
  \label{eq:rj}
  \{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j.
\end{equation}
As $s'_j$ is sampled accordingly to $P(\cdot|s_j,a_j)$ the constructed datapoints help building a good approximation of $R^C(s=s_j,a=a_j)$. A regressor (a simple least-square approximator can do but other solutions such as Gaussian Processes \cite{rasmussen2006gaussian} could also be envisionned) is then fed the datapoints $((s_j,a_i),\hat r_j)$ to obtain $\hat R^C$, a generalisation of $\{((s_j,a_j),\hat r_j)_j\}$ over the whole state-action space. The complete algorithm is given in Alg.~\ref{algo:cascading}.
\begin{algorithm}%[H]
    %\small
  \caption{CSI algorithm}
  \label{algo:cascading}
  \emph{\textbf{Given}} a training set $D_C=\{(s_i,a_i=\pi^E(s_i))\}_{1\leq i \leq D}$ and another training set $D_R=\{(s_{j},a_{j},s'_{j})\}_{1\leq j \leq D'}$\;\\
  \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$\;\\
  \emph{\textbf{Learn}} a reward function $\hat R^C$ from the dataset $\{((s_{j},a_{j}),\hat{r}_j)\}_{1\leq j \leq D'}$, $\forall (s_j,a_j,s'_j) \in D_R,\hat{r}_j=q(s_{j},a_{j})-\gamma q(s'_{j},\pi_C(s'_{j}))$\;\\
  \emph{\textbf{Output}} the reward function $\hat R^{C}$ \;
\end{algorithm}

@@

  The function $q$ is the action-value function for $\pi_C$ with respect to the reward $R^C$, by definition (see Eq.~\eqref{eq:rc}).

@@

More precisely, the expert controls the system for $n$ runs of length $n$, with values of $n$ in $\{3,7,10,15,20\}$, giving $n^2$ samples $(s_k,a_k=\pi_E(a_k),s'_k)_k = D_E$. The dataset $D_C$ fed to the classifier is straightforwardly constructed from $D_E$ by dropping the $s'_k$ terms:
\begin{equation}
  D_C = \{(s_i = s_k,a_i=a_k)_i\}.
\end{equation}
There is no particuliar problem for the classification step. It is however not reasonable to construct the dataset $D_R$ only from expert transitions and expect a small regression error term $\epsilon_R$. Indeed, the dataset $D_E$ only sample the dynamics induced by the expert's policy and not the whole dynamics of the MDP. The regression step also needs samples $((s_j,a_j),\hat r_j)_j$ drawn from non expert actions. We address this problem by using a heuristics.
\subsubsection{Heuristics}
\label{subsubsec:heuristics}
From the raw expert dataset $D_E$ we can construct samples $((s_j=s_k,a_j=a_k),\hat r_j)$ with $\hat r_j$ as specified in Eq.~\eqref{eq:rj}. To compensate for the lack of information about the dynamics of the whole MDP, we assert that disagreeing with the expert's choice is undesirable. Mathematically, we do that by artificially introducting samples $((s_j=s_k,a),r_{min})_{j;\forall a\neq \pi_E(s_j) = a_k}$ where
\begin{equation}
  r_{min} = \min_j\hat r_j - 1.
\end{equation}
Although this heuristics was not analyzed in Sec.~ \ref{sec:analysis} (where the availability of a more complete dataset $D_R$ was assumed), the results shown in the next two subsections demonstrate its soundness.
\subsubsection{Subroutines choices}
\label{subsubsec:subroutines}
The cascading algorithm can be instanciated with some standard classification algorithms and any regression algorithm. The choice of such subroutines may be dictated by the kind and amount of available data, by ease of use or by computational complexity, for example.
